Information in recombination junctions can both compress and improve inferred tree sequences

### 2024-01-22 paper/writeup.tex
discrepancy function write ups
Still under construction, but some progess

\par Now with an algorithm in place
 we need a metric to discern
 if trees modified with extend edges
 in fact correctly infer haplotypes in our original tree sequence.
 One such function a so called \textit{discrepancy function}, can be called on a pair of tree sequences.
 To count all the discrepancies we must begin by counting differences for each node in the tree sequence.

\comment{Note this is somewhat similar to the Robinson-Foulds metric in concept.}

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        Something here \comment{something useful}
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

 Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
 Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
 We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
 Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
 we can measure how much they agree along the genome.
 We do so with the following function
 $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
 where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
 We see $\rho$ measures the fraction of the tree sequence
 along which $x$ and $y$ have exactly the same descendant set. 
 For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
 This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
 This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
 We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
 $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
 Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
 The function $\hat{d}$ measures the proportion of the genome 
 over which the descendant set of $x$ differs from its best matching node $y$.
 We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
 $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
 This function, as we will see in the algorithm below,
 is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.

@@

 \Fn{Shared Spans($T_1,T_2$)}{
 \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
 		Define $a_{ij}=|L_i\cap L_j|$.
 	}
 	\Else{Define $a_{ij}=0$.}
 Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
 }
 \KwRet{$A$}
 }
\BlankLine
\Fn{Discrepancy ($T_1,T_2$)}{
\For{each node $i\in N_{T_1}$}{
	Find nodes in $N_{T_2}$ which have the highest matching span.\;
	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
	Define $\alpha_i = a_{ij}$.\;
}
\begin{left} $$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
\BlankLine\BlankLine
\begin{left} $$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
\BlankLine\BlankLine
\KwRet{tree discrepancy, rmse}
}
\end{algorithm2e}

@@

\begin{figure}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}

>>>>

\par \comment{This needs work. Choose $n_1$ or $i$ for node not both. Do you want it through the perspective of TS or through nodes? You wording is mixed on this}
 Now with an algorithm in place
 we need a metric to discern
 if trees modified with extend edges
 in fact correctly infer haplotypes in our original tree sequence.
 One such function a so called \textit{discrepancy function} (see Algorithm \ref{alg:disc}), can be called on a pair of tree sequences $\T_1$, $\T_2$ of equal length $L$.
 Let $N_1$ and $N_2$ be the set of nodes on $\T_1$ respectively $\T_2$.
 For a pair of nodes $(n_1,n_2)\in N_1\times N_2$, we define their \textit{discrepancy} as a difference in their total spans over the genome.
 Each node $n\in\T$ has a given span along the genome and a time, 
 so we can write each node $n = (t_n, L_n)$ where $L_n\subset \left[0,L\right)$ and $t_n\in \R^{\geq 0}$.
 Clearly, computing the discrepancy between a random pair of nodes is not helpful, 
 so we must first identify which pairs of nodes would be most identical to one another or a 'best match'.
 When we find 'best matching' nodes we want our pair to be close in terms of time and have equal descendant sample sets,
 ie. $s(n_1) = s(n_2)$ where $s$ is the function to compute the descendant sample set of $n_1$.

 To find a best match for each node $n_1\in N_1$ \comment{(Should I do this by mentioning clades?)}
 we compute the \textit{shared span} between $(n_1, m)$ for all $m\in N_2$.
 The \textit{shared span} between two tree sequences, $ss(\T_1,\T_2)$, is an $|N_1|\times |N_2|$ matrix $A$
 where each entry $a_{ij}$ represents the shared span between a pair of nodes $(i,j)\in N_1\times N_2$.
 If a pair has equal descendant sample sets, $s(i) = s(j)$
 then $a_{ij} = |L_i \cap L_j|$.
 If the pair does not have an equal descendant sample set then we set $a_{ij}=0$.
 Matrix $A$ then determines the \textit{best match} for every node $i\in N_1$
 via looking at the maximum in row $i$, $\alpha_i = \max_{j\in N_2}a_{ij}$.
 If this maximum is not unique, we use time difference to determine the best match pair.
 We then define the \textit{discrepancy} between $\T_1$ and $\T_2$ as 
 one minus the proportion of sum of the shared spans divided by the total node span of $\T_1$.
 This function is similar in idea to the Robinson-Foulds metric \comment{CITE}.
 % Robinson-Foulds Paper: https://www.sciencedirect.com/science/article/abs/pii/0025556481900432?via%3Dihub 
 However, where the Robinson-Foulds metric compares nodes between individual pairs of trees,
 in an effort to infer coalescent haplotypes, our function chooses to compare nodes
 across the entire genome. 
 We think this helps track ancestry \comment{something something something...}

\comment{Do we want a complete walkthrough of Figure 3? Could be useful in an appendix}

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \textit{discrepancy function}, $d(T1,T2)$, matches nodes in $T1$ with nodes in $T2$
        based on identical sample sets. We then measure each pair of nodes' \textit{shared span} across the entire tree sequence.
        In this example we have nodes $0$ through $7$ in $T1$ match with their identical counterpart. 
        Node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left{1,2,3\right}$.
        \comment{make some reference to Discrepancy function section or appendix for more in depth calculation.}
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

% Let $TS^1=(\{T_i^1\},N^1,E^1,L^1)$ and $TS^2=(\{T^2_j\},N^2,E^2,L^2)$ be tree sequences.
% Suppose that $x\in N^1$ and $y\in N^2$ are nodes in each tree sequence.
% We denote the \textit{set of descendant samples} of $x$ at position $i$ in the genome as $D_{TS^1}(x,i)$.
% Now for a pair of nodes $(x,y)\in TS^1\times TS^2$ 
% we can measure how much they agree along the genome.
% We do so with the following function
% $$\rho_{(TS^1,TS^2)}(x,y)=\frac{1}{L}\sum_{i=0}^{L-1}[D_{TS^1}(x,i)=D_{TS^2}(y,i)],$$
% where $L=\min\{L^1,L^2\}$ and $[\cdot]$ is the characteristic function. 
% We see $\rho$ measures the fraction of the tree sequence
% along which $x$ and $y$ have exactly the same descendant set. 
% For each node $x\in N^1$ we can use $\rho$ to find its global best matching node $y\in N^2$. 
% This node $y$ should not just be a maximum of $\rho(x,\cdot)$ but it should also have a branch length similar to $x$. 
% This means our node $y$ should minimize both $|t_x-t_y|$ and $1-\rho_{(TS^1,TS^2)}$.
% We then use this $y$ to find count the length of genome which $x$ differs from its best matching node in $TS_2$ with the function
% $$\hat{d}(x,TS^1,TS^2)=\min_y\{|t_x-t_y|(1-\rho_{(TS^1,TS^2)}(x,y))\}.$$
% Here $t_x$ and $t_y$ denote the time of occurrences (ie. branch length) for nodes $x$ and $y$. 
% The function $\hat{d}$ measures the proportion of the genome 
% over which the descendant set of $x$ differs from its best matching node $y$.
% We can now extend this from any node $x\in N^11$ to all of $TS_1$ with the \textbf{discrepancy function}
% $$d(TS_1,TS_2)=\sum_{x\in N_1}\hat{d}(x,TS_1,TS_2).$$
% This function, as we will see in the algorithm below,
% is $|N^1||\{T^1\}|$ as for each tree in $TS_1$ and each node $x\in N^1$ we only need to look at $O(1)$ nodes rather than all nodes in $N^2$.
%

@@

 \Fn{Shared Spans($T_1,T_2$)}{
 \For{pairs $(i,j)\in N_{T_1}\times N_{T_2}$}{
 	$i = (t_i, L_i)$ and $j = (t_j, L_j)$ where $L_i, L_j\subset \left[0,L\right).\quad$ \Comment{Each node is a tuple (time, genome location).}
 	\If{$(i,j)$ has equal sample sets, $s(i)=s(j)$}{
 		Define $a_{ij}=|L_i\cap L_j|$.
 	}
 	\Else{Define $a_{ij}=0$.}
 Define matrix $A=\left[a_{ij}\right]\in M_{N_{T_1}\times N_{T_2}}(\R)$.\;
 }
 \KwRet{$A$}
 }
\BlankLine
\Fn{Discrepancy ($T_1,T_2$)}{
\For{each node $i\in N_{T_1}$}{
	Find nodes in $N_{T_2}$ which have the highest matching span.\;
	$\operatorname{matches}_i = \argmax_{k\in N_{T_2}} a_{ik}.$\;
	Let $j = \argmax_{k\in \operatorname{matches}_i}\frac{1}{1+|t_i-t_k|}$.\;
	Define $\alpha_i = a_{ij}$.\;
	\BlankLine
}
\begin{left}
$$tree\ discrepancy = 1 - \frac{\sum_{i\in N_{T_1}} \alpha_i}{\operatorname{tr}\left(Shared Spans(T_1, T_2)\right)}$$\end{left}\;
\BlankLine\BlankLine
\begin{left}
$$rmse = \sqrt{\frac{\sum_{i\in N_{T_1}}|t_i - t_j|\alpha_i}{\operatorname{tr}\left(Shared Spans(T_1,T_2)\right)}}$$\end{left}\;
\BlankLine\BlankLine
\KwRet{tree discrepancy, rmse}
}
\end{algorithm2e}

@@

\begin{figure}
	\includegraphics[width=0.9\textwidth]{Edge_reduc_comp_speed.pdf}
    \caption{
        Two-part figure, showing reduction in number of edges (left) and increase in speed of stats comptuation (right)
        as a function of number of trees (modulated by sequence length),
        on reasonably big simulated sequences.
        For "speed": plot ratio of runtime before/after extending.
        \label{fig:speed_and_edges}
    }
\end{figure}
